Supplementary material for -- RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-07-17 manuscripts/supplement.Rmd
update supplement with more text describing the SPARQL example #73, doesn't run because of error in newick parsing now

Extract the RDF graph from the nexml

```{r}
rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
graph <- load.rdf("rdf_meta.xml")
```

fetch the NCBI URI for the taxon that has rank 'Order', i.e. the root of the primates. The dot operator `.` between clauses implies a join, in this case

```{r}
root <- sparql.rdf(graph, 
    "SELECT ?uri WHERE { 
        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
    }"               
)
```

Define a function to get the name

 define a recursive function to build newick

        # the newick can be made to contain interior node labels by inserting get_name(id), before the sep="" argument
        return(paste("(", 
                     paste(sapply(children, recurse), 
                           sep = ",", collapse = "," ), 
                     ")",  sep = "", collapse = ""))
    }
}
```

build the tree and visualize it

@@

@@

<!--  Find more consise way to make sure dependencies are cited. 
      How about citing packages that are only on the suggests list? --> 
<!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
<!-- Reference any installation and configuration instructions necessary (e.g. API credentials for figshare) -->

@@

```{r cleanup, include=FALSE, cache=FALSE}
write.bibtex(file="components/sup-references.bib")
unlink("simmap.xml")
unlink("birds.xml")
unlink("geospiza.xml")
unlink("rdf_meta.xml")
```

>>>>

```{r include=FALSE}
library("RNeXML")
```

@@

Rich, semantically meaningful metadata lies at the heart of the NeXML
standard.  R provides a rich environment to unlock this information.
While our previous examples have relied on the user knowing exactly
what metadata they intend to extract (title, publication date, citation
information, and so forth), _semantic_ metadata has meaning that a
computer can make use of, allowing us to make much more conceptually
rich queries than those simple examples.  The SPARQL query language
is a powerful way to make use of such semantic information in making 
complex queries.

While users should consult a formal introduction to SPARQL for further
background, here we illustrate how SPARQL can be used in combination
with R functions in ways that would be much more tedious to assemble
with only traditional/non-semantic queries.  The SPARQL query language
is provided for the R environment through the `rrdf` package `r citep(citation("rrdf"))`,
so we start by loading that package.  We will also make use of functions
from `phytools` and `RNeXML`.

We read in an example file that contains semantic metadata annotations
describing the taxonomic units (OTUs) used in the tree.

In particular, this example declares the taxon rank, NCBI identifier and parent taxon
for each OTU, such as:

```xml
<otu about="#ou541" id="ou541" label="Alouatta guariba">
      <meta href="http://ncbi.nlm.nih.gov/taxonomy/182256" 
            id="ma20" 
            rel="concept:toTaxon" 
            xsi:type="nex:ResourceMeta"/>
      <meta href="http://rs.tdwg.org/ontology/voc/TaxonRank#Species" 
            id="ma21" 
            rel="concept:rank" 
            xsi:type="nex:ResourceMeta"/>
      <meta href="http://ncbi.nlm.nih.gov/taxonomy/9499" 
            id="ma22" 
            rel="rdfs:subClassOf" 
            xsi:type="nex:ResourceMeta"/>
    </otu>

```

In this example, we will construct a cladogram by using this information
to identify the taxonomic rank of each OTU, and its shared parent
taxonomic rank.  (If this example looks complex, try writing down the
steps to do this without the aid of the SPARQL queries).  These examples
show the manipulation of semantic triples, Unique Resource Identifiers
(URIs) and use of the SPARQL "Join" operator.

Note that this example can be run using `demo("sparql", "RNeXML")` to see the 
code displayed in the R terminal and to avoid character errors that
can occur in having to copy and paste from PDF files.

We begin by extracting the RDF graph from the NeXML,

```{r}
rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
tmp <- tempfile()  # rrdf requires a file name, so we must write the XML out first
saveXML(rdf, tmp) 
graph <- load.rdf(tmp)
```

We then fetch the NCBI URI for the taxon that has rank 'Order', i.e. the
root of the primates phylogeny. The dot operator `.` between clauses
implies a join, in this case

```{r}
root <- sparql.rdf(graph, 
"SELECT ?uri WHERE { 
    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
}")
```

This makes use of the SPARQL query language provided by the `rrdf`
package.  We will also define some helper functions that use SPARQL
queries.  Here we define a function to get the name

Next, we define a recursive function to build a newick tree from the taxonomic rank information.

        return(paste("(", 
                     paste(sapply(children, recurse), 
                           sep = ",", collapse = "," ), 
                     ")",  
                     get_name(id), # label interior nodes
                     sep = "", collapse = ""))
    }
}

```

With these functions in place, it is straight forward to build the tree
from the semantic RDFa data and then visualize it

<!-- hoping Rutger can add something here so that read.newick works when we have interior tip labels. -->

@@

@@

```{r cleanup, include=FALSE, cache=FALSE}
write.bibtex(file="components/sup-references.bib")
unlink("simmap.xml")
unlink("birds.xml")
unlink("geospiza.xml")
unlink(tmp)
```
